### 2019-09-15 ic-journal/multivariate.tex
Minor updates (contraction)
Closed Section 5.

$X$ under this specific $P$, by Modus Ponens (MP) the original proof completes.

@@

completes. Thus it remains to prove that

@@

Now consider (formally) what is a solution $\til P$ of CCS equation. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that the single definition
could be used for the solution of all kind of CCS
equations/contractions. Then, of course ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$ must hold, i.e.
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \vdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be pure process, i.e. having no free
variable. This is usually implicitly assumed in the literature but is
actually a \emph{must} to complete some unique-solution proofs.
At last, like the case of CCS equations, we must assume that, the
set of bound variables (i.e. process constants) must be disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs much
easier, and it can be removed in the future.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes. Furthermore, \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}} implicitly
  implies that the two lists $A$ and $B$ have the same length.}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}

\subsection{Unique solution of equations for $\sim$ (the \multivariate version)}

\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Ps}.
       \HOLConst{LENGTH} \HOLBoundVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \HOLBoundVar{P\sp{\prime}}.
           \ensuremath{[}\HOLBoundVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \HOLBoundVar{P\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
           \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{E\sp{\prime}}.
               \HOLConst{context} \HOLFreeVar{Xs} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{FV} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{E\sp{\prime}}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
               \HOLBoundVar{P\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\HOLBoundVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}}
               \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Qs}. \HOLConst{LENGTH} \HOLBoundVar{Qs} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLBoundVar{Qs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{[}\HOLBoundVar{Qs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLBoundVar{E\sp{\prime}}

\HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{STRONG_EQUIV} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Qs} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{STRONG_EQUIV} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{Qs}\hfill{[strong_unique_solution_thm]}
\end{alltt}

\subsection{Unique solution of rooted contractions (the \multivariate version)}

\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{OBS_contracts} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Qs} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{OBS_contracts} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{Qs}\hfill{[unique_solution_of_rooted_contractions]}
\end{alltt}

>>>>

$X$ under this specific $P$, by Modus Ponens (MP) the original proof
\hl{would have completed.}

@@

\hl{would have completed.} Thus it remains to show that

@@

Now consider (formally) what is a solution $\til P$ of CCS equation. First of
all, the definition should be parametrized on a binary CCS
relation $\R$ such as $\sim$ and $\rcontr$, so that the single definition
could be used for the solution of all kind of CCS
equations/contractions. Then, of course ${\til P} \ \R\  {\til E}\{\til
P/\til X\}$ must hold, i.e.
\begin{equation*}
  \begin{cases}
    &P_1 \ \R\ E_1\{\til P/\til X\} \\
    &P_2 \ \R\ E_2\{\til P/\til X\} \\
    & \vdots \\
    &P_n \ \R\ E_n\{\til P/\til X\}
  \end{cases}
\end{equation*}
Furthermore, each $P_i$ should be pure process, i.e. having no free
variable. This is usually implicitly assumed in the literature but is
actually a \emph{must} to complete some unique-solution proofs.
At last, like the case of CCS equations, we must assume that, the
set of bound variables (i.e. process constants) must be disjoint with
$\til X$. (This disjointness requirement is optional but makes many proofs much
easier, and it can be removed in the future.) Putting all together,
below is the formal definition of a solution of \multivariate CCS
equations:\footnote{If $R$ is a binary relation of CCS processes,
  \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}} is the same binary relation but for lists of
  CCS processes. Furthermore, \HOLinline{\HOLConst{LIST_REL}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{A}\\\;\HOLFreeVar{B}} implicitly
  implies that the two lists $A$ and $B$ have the same length.}
\begin{alltt}
   \HOLTokenTurnstile{} \HOLConst{CCS_solution} \HOLFreeVar{R} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenEquiv{}}
      \HOLConst{ALL_PROC} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}} \HOLConst{EVERY} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{e}. \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{e}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\ensuremath{)} \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenConj{}}
      \HOLConst{LIST_REL} \HOLFreeVar{R} \HOLFreeVar{Ps} \ensuremath{(}\HOLConst{MAP} \ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{Es}\ensuremath{)}\hfill{[CCS_solution_def]}
\end{alltt}
Notice that, in HOL, \HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}\\\;\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenEquiv{}}\\\;\HOLFreeVar{Ps}\\\;\HOLSymConst{\HOLTokenIn{}}\\\;\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}. The form in right side suggest that,
\HOLinline{\HOLConst{CCS_solution}\\\;\HOLFreeVar{R}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}} is actually a set containing all
solutions (of \HOLinline{\HOLConst{CCS_equation}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Es}}). \hlC{Then the unique-solution
theorems can be understood as: any two (syntactically different) elements
in the set are bisimilar.}

\subsection{Unique solution of equations/contractions (the \multivariate version)}

With all above new devices (\multivariate contexts and substitutions),
the \multivariate case of Lemma~\ref{lem:milner313} is formalised
below:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Ps}.
       \HOLConst{LENGTH} \HOLBoundVar{Ps} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}}
       \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \HOLBoundVar{P\sp{\prime}}.
           \ensuremath{[}\HOLBoundVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \HOLBoundVar{P\sp{\prime}} \HOLSymConst{\HOLTokenImp{}}
           \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{E\sp{\prime}}.
               \HOLConst{context} \HOLFreeVar{Xs} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{FV} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{set} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenConj{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{BV} \HOLBoundVar{E\sp{\prime}}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
               \HOLBoundVar{P\sp{\prime}} \HOLSymConst{\ensuremath{=}} \ensuremath{[}\HOLBoundVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLBoundVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}}
               \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{Qs}. \HOLConst{LENGTH} \HOLBoundVar{Qs} \HOLSymConst{\ensuremath{=}} \HOLConst{LENGTH} \HOLFreeVar{Xs} \HOLSymConst{\HOLTokenImp{}} \ensuremath{[}\HOLBoundVar{Qs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLFreeVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{[}\HOLBoundVar{Qs}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]} \HOLBoundVar{E\sp{\prime}}
\end{alltt}
Notice that, while Milner did not pointed out explicitly, the involved
weakly guarded context $E$ must contain free variables up to $\til
X$~--- the same requirement as in Theorem~\ref{t:Mil89s1}, which
otherwise does not complete. Besides, we further required that, the
bound variables of $E$ must be disjoint with $\til X$.

The \multivariate version of Theorem~\ref{t:Mil89s1} is formalised
below:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{STRONG_EQUIV} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Qs} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{STRONG_EQUIV} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{Qs}\hfill{[strong_unique_solution_thm]}
\end{alltt}

And the \multivariate version of Theorem~\ref{t:rcontraBisimulationU}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_equation} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}} \HOLConst{weakly_guarded} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{OBS_contracts} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenConj{}}
   \HOLFreeVar{Qs} \HOLSymConst{\HOLTokenIn{}} \HOLConst{CCS_solution} \HOLConst{OBS_contracts} \HOLFreeVar{Xs} \HOLFreeVar{Es} \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{Ps} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{Qs}\hfill{[unique_solution_of_rooted_contractions]}
\end{alltt}

Compared with their formal proofs in the \univariate case, although each
step related to \multivariate substitutions is more difficult than
the \univariate case, the entire proofs still follow the same outlines,
and many proof scripts can be directly copied from their \univariate version.
